\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-05 coqimpl.tex
coqimpl

% \subsection{Language Design}

% \subsection{Plugin Implementation}

>>>>

\input{stlc-compiled.tex}
\input{stlcfix-compiled.tex}

We implement our language design as a Coq plugin.
Rather than modifying the Coq kernel, a prototypical implementation as a
plugin allows a clearly defined trusted base and
improved compatibility with different versions of Coq.

The implementation works by translating programs in \Lang syntax into
programs that can be checked and evaluated by Coq.
The translation is compatible with interactive theorem proving, in that
a family is translated piece by piece, allowing each field to be defined
and checked separately.
The translation is modular and efficient, in that
code inherited from a base family need not be retranslated or rechecked.

\noindentparagraph{Explicit self parameterization.}

The spirit of the translation is to take ``family polymorphism''
literally: every field is translated into a Coq definition that is
polymorphic to (i.e., universally quantified over) a representation of
its enclosing family.
While this universal quantification has been implicit with the \Lang
syntax, it has to be made explicit in the translated Coq code.

\cref{fig:stlc-compiled,fig:stlcfix-compiled} illustrate the translation of the
\lsti{STLC} and \lsti{STLCFix} families from \cref{fig:stlc-mechanized}.
Fields of a family are translated into parameterized Coq modules
(or parameterized module types).

For instance, consider field \lsti{env} in family \lsti{STLC}.
It is translated into a top-level module named \lsti{STLC\_env}.
This module has a parameter called \lsti{self} representing the enclosing
family: fields of the current family in the context of \lsti{env} can be
referenced through this \lsti{self}.
In particular, \lsti{env} is defined as \lsti{id -> option ty},
where \lsti{ty} is a late-bound reference to the \lsti{ty} field of the
enclosing family.
Hence, this reference to \lsti{ty} is translated in \lsti{STLC\_env} to \lsti{self.ty},
which is manifestly polymorphic to the enclosing family.
This translation of the \lsti{env} field can be shared
with a derived family even if it extends \lsti{ty} (e.g., \lsti{STLCProd})---%
no recompilation is needed because \lsti{self.ty} is not tied to a
particular definition of \lsti{ty}.

The type of \lsti{STLC\_env}'s \lsti{self} parameter is \lsti{STLC\_env\_Ctx},
a module type constructed from the translation of \lsti{env}'s preceding
field, \lsti{STLC\_ty}, and its context \lsti{STLC\_ty\_Ctx}.
In turn, \lsti{STLC\_ty\_Ctx} (not shown in \cref{fig:stlc-compiled}) is
constructed from the translation of \lsti{ty}'s preceding field,
\lsti{STLC\_subst}, and its context \lsti{STLC\_subst\_Ctx}.
Thus, the \lsti{self} parameter can be used to reference those and only
those fields in the current field's typing context, which echoes the
discussion in \cref{sec:sound}.

\noindentparagraph{Translating extensible inductive types.}

An \lsti{FInductive} definition is translated to a parameterized module type.
Consider the inductive type \lsti{tm}.
In \lsti{STLC}, it is translated to a top-level module type
\lsti{STLC\_tm} that declares a \lsti{tm} type,
its constructors (\lsti{tm_unit} etc.),
its partial recursor (\lsti{tm_prect_STLC}),
and the propositional equalities on the partial recursor (\lsti{tm_unit_eq_STLC} etc.).

Importantly,
\lsti{STLC\_tm} merely declares the existence of these names and their types;
it does not specify their definitions or proofs.
Knowing the existence of these names and their types
suffices to type-check and translate the rest of the family.
Leaving the definitions and proofs open enables \lsti{STLC} and \lsti{STLCFix}
to instantiate \lsti{tm} differently upon \lsti{End}~\lsti{STLC} and \lsti{End}~\lsti{STLCFix}.

The command \lsti{FInductive tm : Set += tm_fix : }\dadada~in family
\lsti{STLCFix} is again translated to a module type \lsti{STLCFix\_tm} (\cref{fig:stlcfix-compiled}).
It includes all names declared by \lsti{STLC\_tm}, and additionally declares
the new constructor \lsti{tm_fix} and related equalities.

\noindentparagraph{Translating recursion and induction.}

An \lsti{FRecursion} definition is translated in two parts:
first a module containing the definitions of all the case handlers,
and then a module type declaring the existence of the recursive function as well as
its propositional equalities.

Consider the translation of \lsti{subst} in family \lsti{STLC}.
First, a module named \lsti{STLC\_subst\_Cases} is generated interactively:
every time the programmer completes a \lsti{Case}, a case handler (e.g., \lsti{Def subst\_tm_unit})
is generated and added to the module.

Upon \lsti{End}~\lsti{subst}, a module type named \lsti{STLC\_subst} is generated.
%
As discussed in \cref{sec:latebind}, \lsti{subst} can be further bound,
so its definition is not exposed to the succeeding fields.
Accordingly, the translation \lsti{STLC\_subst} merely declares the type of \lsti{subst}
and the existence of its propositional equalities, leaving \lsti{subst} undefined
and the equalities unproven.
%
The propositional equalities are stated in terms of the case handlers,
whose definitions \emph{are} available through the \lsti{self} parameter.
For example, the type of \lsti{subst_tm_unit_eq} can be simplified to
\lsti{\forall x t, self.subst self.tm_unit x t = self.tm_unit}.
The equalities will be available for use in the checking and translation of
the succeeding fields through their \lsti{self} parameters.

Importantly, code generated for the case handlers is shared with derived families.
In \cref{fig:stlcfix-compiled}, \lsti{STLCFix\_subst\_Cases}
reuses---without rechecking---all the case handlers in
\lsti{STLC\_subst\_Cases} via the command \lsti{Include STLC\_subst\_Cases(self)}.

The translation of \lsti{FInduction} is similar, except that there is no need to
register propositional equalities, as \lsti{FInduction} proofs are considered opaque.

\noindentparagraph{Translation of further-bindables vs.\ non-further-bindables.}

In family \lsti{STLC},
field \lsti{env} and the case handlers for \lsti{subst}
are not further-bindable by derived families.
In contrast, \lsti{tm}, \lsti{subst}, and the related equalities
can be further bound.
The distinction is reflected in the translations.
Further-bindable fields are translated to module types that export only types of the fields.
Non-further-bindable fields are translated to modules that export definitional equalities
on the fields.

%\footnotetext{%
%Axioms in Coq module types are more like parameters than real axioms.
%They become part of the signature described by the module type.
%}

\noindentparagraph{Eliminating \lsti{self} by aggregation.}

Upon the conclusion of a family definition (e.g., \lsti{End}~\lsti{STLCFix}),
\lsti{self} parameters in the translations of the fields must be instantiated,
so that references like \lsti{STLCFix.typesafe} can be translated.
This is done by assembling the \lsti{self}-parameterized pieces into a single module.
%
For example, module \lsti{STLC} in \cref{fig:stlc-compiled} is generated
upon \lsti{End}~\lsti{STLCFix}.
All the fields are added to this module in the same order as they appear in the family definition.

For the further-bindables, \lsti{Axiom}s declared in the module types must be instantiated.
%
\begin{itemize}[align=left,itemsep=5pt,labelsep=*,leftmargin=*]

\item
For \lsti{tm},
an inductive type \lsti{tm\_} is generated, with \lsti{tm\_}
instantiating the axiomatized \lsti{tm} type, and with
its constructors (\lsti{tm_unit\_} etc.) instantiating 
the axiomatized constructors (\lsti{tm_unit} etc.).
%
The partial recursor \lsti{tm_prect_STLC} is instantiated
%to \lsti{\lambda}\,\lsti{P, tm\__rect (\lambda}\,\lsti{t, option (P t))},
with the help of \lsti{tm\_rect}, the recursor of the generated \lsti{tm\_} type.
Proving the equalities on \lsti{tm_prect_STLC} is then trivial, by \lsti{reflexivity}.

\item
Similarly, \lsti{subst} is instantiated by applying the recursor \lsti{tm\_rect}
to the (already included) case handlers (\lsti{subst\_tm_unit}, etc.).
Propositional equalities on \lsti{subst} are then trivial, by \lsti{reflexivity}.

\end{itemize}

For the non-further-bindables, the translated modules are directly included
(e.g., \lsti{Include STLC\_subst\_Cases} and \lsti{Include STLC\_env}).
%
The instantiations of \lsti{self}s are implicit thanks to a Coq nicety:
when including a higher-order module, Coq automatically instantiates its
parameter with the current interactive module environment.
For instance, command \lsti{Include STLC\_subst\_Cases} is successfully executed,
because Coq automatically instantiates \lsti{self} using the current
module environment, which contains all the fields required by \lsti{STLC\_subst\_Cases\_Ctx}.

Module \lsti{STLCFix} in \cref{fig:stlcfix-compiled} is generated upon
\lsti{End}~\lsti{STLCFix}, in the same fashion described above for \lsti{STLC}.
The translation makes code and proof sharing evident.
For example, the four case handlers for \lsti{subst} compiled for \lsti{STLC}
are reused by module \lsti{STLCFix}. So is the proof of \lsti{typesafe}.

Finally, the reference \lsti{STLCFix.typesafe} (where \lsti{STLCFix} is a family)
can simply be translated to \lsti{STLCFix.typesafe} (where \lsti{STLCFix} is a module),
as the last line of \cref{fig:stlcfix-compiled} shows.

\ifShowOldWriting

\newpage

### 2022-11-05 stlc-compiled.tex
coqimpl

@@

[@\outdent@]Module STLC\_subst\_Cases (self : STLC\_subst\_Ctx).

@@

@@

Axiom subst_tm_unit_eq : \forall x t,
  self.subst (self.tm_unit) x t = self.tm_unit.

@@

[@\outdent@]End STLC\_subst.[@\tanc{c-STLC-subst-end}@][@\outersep@]
[@\outdent@][@\codecomment{Code emitted upon definition of \texttt{ty} in family \texttt{STLC}}@] [@\dadada@][@\tanc{c-STLC-ty-start}@][@\tanc{c-STLC-ty-end}@][@\outersep@]

@@

[@\outdent@][@\codecomment{Code emitted for other fields defined in family \texttt{STLC}}@] [@\dadada@][@\tanc{c-STLC-rest}@][@\outersep@]
[@\outdent@][@\codecomment{Code emitted upon completion of family \texttt{STLC}}@][@\tanc{c-STLC-start}@]

@@

Def tm := tm\_.[@\hfill\codecomment{Instantiate \texttt{tm} \& its constructors}@]

@@

[@\hfill\codecomment{Instantiate \texttt{tm} partial recursor \& its equalities}@]
Def tm_prect_STLC P := tm\__rect (\lambda[@\,@]t,[@\,@]option[@\,@](P[@\,@]t)).

@@

Def subst[@\,@] :=[@\hfill\codecomment{Instantiate \texttt{subst} \& its equalities}@]
  tm\__rect _ subst\_tm_unit subst\_tm_var

@@

Fact[@\;@][@\textls[-40]{subst\_tm\_unit\_eq}@][@\,@]:[@\,@][@\dadada@].[@\,@][@\textls[-40]{{\color{dkblue}reflexivity}.}@][@\,@]Qed.[@\hfill@]Fact [@\dadada@][@\tanc{c-STLC-d-subst-end}@][@\tanc{c-STLC-firstcol-end}\columnbreak@][@\innersep@]
[@\dadada@][@\hfill\codecomment{Instantiate \texttt{ty}, its constructors, partial recusor \& eq.}@][@\tanc{c-STLC-d-ty}@][@\tanc{c-STLC-sndcol-start}@][@\innersep@]

@@

[@\dadada@][@\hfill\codecomment{Include/Instantiate other fields of family \texttt{STLC}}@][@\tanc{c-STLC-d-rest}@][@\innersep@]

@@

\makeline[-1pt]{c-STLC-subst-start}{c-STLC-subst-end}[codecomment-color!50][1.8pt]
\makeline[-1pt]{c-STLC-tm-start}{c-STLC-tm-end}[codecomment-color!50][1.8pt]
\makeline[-1pt]{c-STLC-ty-start}{c-STLC-ty-end}[codecomment-color!50][1.8pt]
\makeline[-1pt]{c-STLC-env-start}{c-STLC-env-end}[codecomment-color!50][1.8pt]
\makeline[-1pt]{c-STLC-rest}{c-STLC-rest}[codecomment-color!50][1.8pt]
\makeline[-1pt]{c-STLC-start}{c-STLC-firstcol-end}[codecomment-color!50][1.8pt]
\makeline[4pt]{c-STLC-d-tm-start}{c-STLC-d-tm-end}[codecomment-color!35][1.0pt]
\makeline[4pt]{c-STLC-d-subst-start}{c-STLC-firstcol-end}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+8pt]{c-STLC-sndcol-start}{c-STLC-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+12pt]{c-STLC-d-ty}{c-STLC-d-ty}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+12pt]{c-STLC-d-env}{c-STLC-d-env}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+12pt]{c-STLC-d-rest}{c-STLC-d-rest}[codecomment-color!35][1.0pt]

@@

\colorlet{codecomment-color}{orange}

\begin{lstlisting}
[@\outdent@][@\codecomment{Code emitted upon definition of \texttt{tm} in family \texttt{STLCFix}}@][@\tanc{c-STLCFix-tm-start}@]
[@\outdent@]Module Type STLCFix\_tm\_Ctx.
[@\outdent@]End STLCFix\_tm\_Ctx.[@\innersep@]
[@\outdent@]Module Type STLCFix\_tm (self : STLCFix\_tm\_Ctx).
Include STLC\_tm(self).
Axiom tm_fix : id -> tm -> tm.
Axiom tm_fix_eq_STLC : \forall[@\,@][@\dadada@],[@\,@]tm_prect_STLC [@\dadada@][@\;@]=[@\;@]None.
Axiom tm_prect_STLCFix : [@\dadada@].
Axiom tm_fix_eq_STLCFix : [@\dadada@].
[@\outdent@]End STLCFix\_tm.[@\tanc{c-STLCFix-tm-end}@][@\outersep@]
[@\outdent@][@\codecomment{Code emitted upon definition of \texttt{subst} in family \texttt{STLCFix}}@][@\tanc{c-STLCFix-subst-start}@]
[@\outdent@]Module Type STLCFix\_subst\_Ctx.
Include STLCFix\_tm\_Ctx. Include STLCFix\_tm.
[@\outdent@]End STLCFix\_subst\_Ctx.[@\innersep@]
[@\outdent@]Module STLCFix\_subst\_Cases (self[@\,@]:[@\,@]STLCFix\_subst\_Ctx).
Include STLC\_subst\_Cases(self).
Def subst\_tm_fix := [@\dadada@].
[@\outdent@]End STLCFix\_subst\_Cases.[@\innersep@]
[@\outdent@]Module Type STLCFix\_subst (self[@\,@]:[@\,@]STLCFix\_subst\_Ctx).
Include STLC\_subst(self).
Axiom subst_tm_fix_eq : [@\dadada@].
[@\outdent@]End STLCFix\_subst.[@\tanc{c-STLCFix-subst-end}@][@\outersep@]
[@\outdent@][@\codecomment{Code emitted for other fields defined in family \texttt{STLCFix}}@] [@\dadada@][@\tanc{c-STLCFix-rest}@][@\outersep@]
[@\outdent\codecomment{Code emitted upon completion of family \texttt{STLCFix}}@][@\tanc{c-STLCFix-start}@]
[@\outdent@]Module STLCFix.[@\innersep@]
Inductive tm\_ : Set :=[@\tanc{c-STLCFix-d-tm-start}@]
| tm_unit\_[@\,@]:[@\,@]tm\_ | tm_var\_[@\,@]:[@\,@]id -> tm\_
| tm_abs\_[@\,@]:[@\,@]id->tm\_->tm\_ | tm_app\_[@\,@]:[@\,@]tm\_->tm\_->tm\_
| tm_fix\_[@\,@]:[@\,@]id->tm\_->tm\_.
Def tm := tm\_.[@\hfill\codecomment{Instantiate \texttt{tm} \& its constructors}@]
Def tm_unit := tm_unit\_. Def tm_var := tm_var\_.
Def tm_abs  := tm_abs\_.  Def tm_app := tm_app\_.
Def tm_fix  := tm_fix\_.
[@\dadada@][@\hfill\codecomment{Instantiate \texttt{tm} partial recursors \& their equalities}@][@\tanc{c-STLCFix-d-tm-end}@][@\innersep@]
Include STLCFix\_subst\_Cases. [@\tanc{c-STLCFix-d-subst-start}@]
Def subst :=[@\hfill\codecomment{Instantiate \texttt{subst} \& its equalities}@]
  tm\__rect _ subst\_tm_unit subst\_tm_var
             subst\_tm_abs  subst\_tm_app subst\_tm_fix.
Fact[@\;@][@\textls[-40]{subst\_tm\_unit\_eq}@][@\,@]:[@\,@][@\dadada@].[@\,@][@\textls[-40]{{\color{dkblue}reflexivity}.}@][@\,@]Qed.[@\hfill@]Fact [@\dadada@][@\tanc{c-STLCFix-d-subst-end}@][@\innersep@]
[@\dadada@][@\hfill\codecomment{\textls[-10]{Instantiate \texttt{ty}, its constructors, partial recursor \& eq.}}@][@\tanc{c-STLCFix-d-ty}@][@\innersep@]
Include STLC\_env.[@\hfill\codecomment{Include \texttt{env}}@][@\tanc{c-STLCFix-d-env}@][@\innersep@]
[@\dadada@][@\hfill\codecomment{Include/Instantiate other fields of family \texttt{STLCFix}}@][@\tanc{c-STLCFix-d-rest}@][@\innersep@]
[@\outdent@]End STLCFix.[@\tanc{c-STLCFix-end}@]
\end{lstlisting}

\makeline[.5\textwidth+8pt]{c-STLCFix-tm-start}{c-STLCFix-tm-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+8pt]{c-STLCFix-subst-start}{c-STLCFix-subst-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+8pt]{c-STLCFix-rest}{c-STLCFix-rest}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+8pt]{c-STLCFix-start}{c-STLCFix-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+13pt]{c-STLCFix-d-tm-start}{c-STLCFix-d-tm-end}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+13pt]{c-STLCFix-d-subst-start}{c-STLCFix-d-subst-end}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+13pt]{c-STLCFix-d-ty}{c-STLCFix-d-ty}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+13pt]{c-STLCFix-d-env}{c-STLCFix-d-env}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+13pt]{c-STLCFix-d-rest}{c-STLCFix-d-rest}[codecomment-color!35][1.0pt]

\definecolor{codecomment-color}{HTML}{000080}

\begin{lstlisting}
[@\outdent@][@\codecomment{Code emitted upon command \texttt{\textls[-35]{Check STLCFix.typesafe}}}@][@\tanc{c-Check-STLCFix-typesafe-start}@]
[@\outdent@]Check STLCFix.typesafe.[@\tanc{c-Check-STLCFix-typesafe-end}@]
\end{lstlisting}

\makeline[.5\textwidth+8pt]{c-Check-STLCFix-typesafe-start}{c-Check-STLCFix-typesafe-end}[codecomment-color!50][1.8pt]

%\vspace{-10pt}

\caption{Compilation of the mechanization of the STLCs in \cref{fig:stlc-mechanized} to parameterized Coq modules.}
\label{fig:stlc-compiled}
\end{figure}

>>>>

[@\outdent@]Module Type STLC\_subst\_Cases\_Ctx.

@@

[@\outdent@]End STLC\_subst\_Cases\_Ctx.[@\innersep@]
[@\outdent@]Module [@\textls[-30]{STLC}@]\_[@\textls[-30]{subst}@]\_[@\textls[-30]{Cases}@] (self[@\,@]:[@\,@][@\textls[-30]{STLC}@]\_[@\textls[-30]{subst}@]\_[@\textls[-30]{Cases}@]\_[@\textls[-30]{Ctx}@]).

@@

Include STLC\_subst\_Cases\_Ctx.
Include STLC\_subst\_Cases.

@@

Axiom subst_tm_unit_eq :
  \forall x t, self.subst (self.tm_unit) x t =
  self.subst\_tm_unit x t.

@@

[@\outdent@]End STLC\_subst.[@\tanc{c-STLC-subst-end}@][@\columnbreak@]
[@\outdent@][@\codecomment{Code emitted upon definition of \texttt{ty} in family \texttt{STLC}}@] [@\tanc{c-STLC-ty-start}@]
[@\dadada@][@\tanc{c-STLC-ty-end}@][@\outersep@]

@@

[@\outdent@][@\codecomment{Code emitted for other fields defined in family \texttt{STLC}}@][@\tanc{c-STLC-rest-start}@]
[@\dadada@][@\tanc{c-STLC-rest-end}@][@\outersep@]
[@\outdent@][@\codecomment{Code emitted upon conclusion of family \texttt{STLC}}@][@\tanc{c-STLC-start}@]

@@

[@\codecomment{Instantiate \texttt{tm} \& its constructors}@]
Def tm := tm\_.

@@

[@\codecomment{Instantiate \texttt{tm} partial recursor \& its equalities}@]
Def tm_prect_STLC := \lambda P, tm\__rect (\lambda[@\,@]t, option (P t)).

@@

[@\codecomment{Instantiate \texttt{subst} \& its equalities}@]
Def subst[@\,@] := tm\__rect _ subst\_tm_unit subst\_tm_var

@@

Fact[@\;@][@\textls[-40]{subst\_tm\_unit\_eq}@][@\,@]:[@\,@][@\dadada@].[@\,@][@\textls[-40]{{\color{dkblue}reflexivity}.}@][@\,@]Qed.[@\hfill@]Fact [@\dadada@][@\tanc{c-STLC-d-subst-end}@][@\tanc{c-STLC-firstcol-end}@][@\innersep@]
[@\codecomment{Instantiate \texttt{ty}, its constructors, partial recusor \& eq.}@] [@\dadada@][@\tanc{c-STLC-d-ty}@][@\tanc{c-STLC-sndcol-start}@][@\innersep@]

@@

[@\codecomment{Include/Instantiate other fields of family \texttt{STLC}}@] [@\dadada@][@\tanc{c-STLC-d-rest}@][@\innersep@]

@@

\makeline[-1pt]{c-STLC-tm-start}{c-STLC-tm-end}[codecomment-color!50][1.8pt]
\makeline[-1pt]{c-STLC-subst-start}{c-STLC-subst-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+8pt]{c-STLC-ty-start}{c-STLC-ty-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+8pt]{c-STLC-env-start}{c-STLC-env-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+8pt]{c-STLC-rest-start}{c-STLC-rest-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+8pt]{c-STLC-start}{c-STLC-end}[codecomment-color!50][1.8pt]
\makeline[.5\textwidth+12pt]{c-STLC-d-tm-start}{c-STLC-d-tm-end}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+12pt]{c-STLC-d-subst-start}{c-STLC-d-subst-end}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+12pt]{c-STLC-d-ty}{c-STLC-d-ty}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+12pt]{c-STLC-d-env}{c-STLC-d-env}[codecomment-color!35][1.0pt]
\makeline[.5\textwidth+12pt]{c-STLC-d-rest}{c-STLC-d-rest}[codecomment-color!35][1.0pt]

@@

\caption{Compilation of family \lsti!STLC! (\cref{fig:stlc-mechanized}).}
\label{fig:stlc-compiled}
\end{figure}
